---
title: "Data Tidying"
author: "Danica Schaffer-Smith"
date: "11/6/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load pkgs, warning=F, message=F}
library(dplyr)
library(tidyr)
```

# Read and Clean data

```{r}
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", 
                               method = "libcurl"),
                    stringsAsFactors = FALSE)
head(catch_original)
```

Remove `all` column and `notes` column because we don't need them.
```{r}
catch_data <- catch_original %>% 
  select(-All, -notesRegCode)
  #select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
```

Quick QA check:
```{r}
summary(catch_data) # Chinook is a character!  Why!?
```

## Fix Chinook column

Use `mutate` to fix Chinook column. First, overwrite a bad value that was read in as "I" instead of 1 by OCR. 
```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook))

summary(catch_clean)
```

Finding the rows that got turned into NA.
```{r}
i <- which(is.na(catch_clean$Chinook))
i
```

```{r}
catch_original[i, ] # Look at that record [rows, columns]
```

# Reshape Data

```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to = "count")
head(catch_long)
```

```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = Year,
              values_from = count)
head(catch_wide)
```
`rename` count column to `catch_thousands`

```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = count) %>% 
  mutate(catch = catch_thousands * 1000) %>% 
  select(-catch_thousands)

head(catch_long)
```


# Summarize Data

Use group by and summarize to create a summary table for our dataset. 
```{r}
mean_region <- catch_long %>% 
  group_by(Region, species) %>% 
  summarize(catch_mean = mean(catch),
            num_obs = n())

mean_region
```

Calculate the total catch by species. Which is the most commonly caught across all years and regions?

```{r}
sum_long <- catch_long %>% 
  group_by(species) %>% 
  summarize(catch_sum = sum(catch)) %>%
  arrange(-catch_sum)
  #arrange(desc(catch_sum))
sum_long
```

Filter to examine years pre-1900. The most dominant species was Sockeye!
```{r}
sum_pre1900 <- catch_long %>% 
  filter(Year<1900) %>% 
  group_by(species) %>% 
  summarise(total = sum(catch)) %>% 
  arrange(desc(total))
sum_pre1900
```


# Join Data 

We can use joins to bring multiple tables together. We have a regions dataframe for the salmon catch data that can help us to better describe regions. We want to keep all of our observations.

First, read in the regions dataset

```{r}
region_defs <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", method = "libcurl"), stringsAsFactors = F) %>% 
  select(code, mgmtArea) %>% 
  rename(Region = code)

head(region_defs)
```

Join the dataframes together
```{r}
#catch_joined <- left_join(catch_long, region_defs, by = c("Region"="code"))

catch_joined <- left_join(catch_long, region_defs, by = "Region")

head(catch_joined)
```

# Separate and Unite

```{r}
# This is a made up example. The samples were observed with Islands (first three letters) and site number
site_codes <- data.frame(site = c("HAW-100",
                                  "HAW-101",
                                  "OAH-102",
                                  "OAH-103",
                                  "MAI-100"),
                         stringsAsFactors = F)
site_codes
```

Separate island from site number using `separate`

```{r}
site_codes_split <- site_codes %>% 
  #separate(site, into = c("island", "site_num"), sep = "-")
  separate(site, into = c("island", "site_num"), sep = "-", remove = F)
site_codes_split
```

Reunited and it feels so good.

```{r}
united <- site_codes_split %>% 
  unite(site_code, island, site_num, sep = "_") #name of the new column, name of 1st variable, name of 2nd variable, separator. 

united
```

Always knit your document to make sure it runs!

You probably want to clean it up a bit and provide some more information for other users. 